Sharing and organizing research products as R packages

### 2020-05-18 manuscript/manuscript.Rmd
Add Author Contributions section to MS
Revision draft 1

Research projects produce experiments, data, analyses, manuscripts, posters, slides, stimuli and materials, computational models, and more. However, the potential added value of these products is not fully realized due to limited sharing and curating practices. Although more transparent communication of these research products has recently been encouraged [@HoutkoopDataSharingPsychology2018; @LindsaySharingDataMaterials2017; @VanpaemelAreWeWasting2015; @Wichertspooravailabilitypsychological2006; @KleinPracticalGuideTransparency2018; @MartoneDatasharingpsychology2018; @RouderMinimizingMistakesPsychological2019; @Rouderwhatwhyhow2016], these efforts often focus narrowly on sharing data (and sometimes analysis code). Further, the practical value of sharing is often limited by poor documentation, incompatible file formats, and lack of organization, resulting in low rates of reproducibility [@HardwickeDataavailabilityreusability2018]. Standardization of protocols for sharing would be beneficial; but, such standards have not emerged, possibly due to the variance of research in psychology. Instead of developing another standard, we suggest borrowing existing standards and practices from software engineering. Specifically, the R package standard, with additional R authoring tools, provides a robust framework for organizing and sharing reproducible research products.

Some advances in data-sharing standards have emerged: In much of psychological science it is now customary to share data on Open Science Framework (OSF). However, those materials often contain idiosyncratic file organization and minimal or missing documentation for raw data. In specific areas, organization and documentation standards have emerged, [e.g., the BIDS framework in neuroscience, @gorgolewski_brain_2016], but they usually only consider data and code instead of the project as a whole. More comprehensive proposals are described in the Transparency and Openness Promotion [@nosek_promoting_2015], and Peer Reviewers' Openness initiative guidelines [@morey_peer_2016], but these fall short of describing detailed standards for organization and metadata. Additionally, these standards (if they exist) may not be widely known, recognized, agreed upon, and/or adopted at large.

@@

Learning and following R conventions for declaring functions has a pedagogical benefit to the researcher and may improve their practices. There is also a reuse benefit: Functions can be difficult to find in old scripts, but easy to find and load if they are called from an existing package. Thus, formally including one's functions in R packages facilitates reproducibility and sharing.

@@

First, raw data in any format is stored in the "data-raw" directory to ensure that the method of sharing remains software-agnostic. Second, instructions for converting and cleaning the raw data into analyzable format should also be included in this directory (the raw data files should not themselves be modified). These pre-processing steps should be saved so that the complete path from raw data to results is transparent and reproducible, preferably in an R script in the "data-raw" directory. That script should end with creating an R data object in the "data" directory. Placing the R data object there ensures that it is included in the resulting R package, which makes using the data effortless for any R user, and eschews the need to download additional files. Loading the R package makes the included data immediately available in the R console.

Finally, the resulting data object in "data" should be documented in a standardized format by placing a `data.R` file in the "R" directory, a process discussed in more detail in our complete online supplementary tutorial, and in @WickhamPackagesOrganizeTest2015. That standard format of documentation is especially useful because it allows datasets' documentation to be viewed in R (e.g., type `?attitude` in the R console) and online (see below).

@@

The core of our proposal is that all project materials are stored in one directory on the user's local filesystem. When version control, such as Git, is enabled in this directory, the entire project can be seamlessly shared with others through online platforms such as GitHub, GitLab, and OSF. Version control systems provide filesystems with additional functioning for keeping track of files' versions, improved options for collaborating, and seamless integration with online collaboration platforms; however, the details are outside our scope, and we refer readers to Vuorre & Curley [-@VuorreCuratingResearchAssets2018]. By default, *vertical* projects enable Git for new projects.

@@

If the project contains R functions, data, and documentation following the R package standard, then others can install the package on their computers for fast access to the data and functions from R. Again, integrating the project with GitHub is beneficial, because then others can install the package from GitHub using the *devtools* package, as was shown above in how to install *vertical*. Sharing as an R package is very convenient for other R users, as any packaged data and functions are immediately available for use, natively in R, after package installation. We expand on the extended benefits of using R packages for transparency and reproducibility in the general discussion.

@@

@@

@@

>>>>

Research projects produce experiments, data, analyses, manuscripts, posters, slides, stimuli and materials, computational models, and more. However, the potential added value of these products is not fully realized due to limited sharing and curating practices. Although more transparent communication of these research products has recently been encouraged [@HoutkoopDataSharingPsychology2018; @LindsaySharingDataMaterials2017; @VanpaemelAreWeWasting2015; @Wichertspooravailabilitypsychological2006; @KleinPracticalGuideTransparency2018; @MartoneDatasharingpsychology2018; @RouderMinimizingMistakesPsychological2019; @Rouderwhatwhyhow2016], these efforts often focus narrowly on sharing data (and sometimes analysis code). Further, the practical value of sharing is often limited by poor documentation, incompatible file formats, and lack of organization, resulting in low rates of reproducibility[^reproducibility] [@HardwickeDataavailabilityreusability2018]. Standardization of protocols for sharing would be beneficial; but, such standards have not emerged, possibly due to the variance of research in psychology. Instead of developing another standard, we suggest borrowing existing standards and practices from software engineering. Specifically, the R package standard, with additional R authoring tools, provides a robust framework for organizing and sharing reproducible research products.

[^reproducibility]: By "reproducibility", we mean computational or analytic reproducibility by which previous results can be reproduced using existing data. Contrast this to "replicability", in which entire procedures are repeated to generate new data to test a scientific idea.

Some advances in data-sharing standards have emerged: It is becoming more popular to share data on the Open Science Framework (OSF). However, those materials often contain idiosyncratic file organization and minimal or missing documentation for raw data. In specific areas, organization and documentation standards have emerged, [e.g., the BIDS framework in neuroscience, @gorgolewski_brain_2016], but they usually only consider data and code instead of the project as a whole. More comprehensive proposals are described in the Transparency and Openness Promotion [@nosek_promoting_2015], and Peer Reviewers' Openness initiative guidelines [@morey_peer_2016], but these fall short of describing detailed standards for organization and metadata. Additionally, these standards (if they exist) may not be widely known, recognized, agreed upon, and/or adopted at large.

@@

Wrapping packages also provide a reuse benefit: Functions can be difficult to find in old scripts, but easy to find and load if they are called from an existing package. Thus, formally including one's functions in R packages facilitates reproducibility and sharing.

@@

First, raw data in any format is stored in the "data-raw" directory to ensure that the method of sharing remains software-agnostic. If possible, that raw data should be in a text based format, rather than proprietary formats such as SPSS files. Second, instructions for converting and cleaning the raw data into analyzable format should also be included in this directory (the raw data files should not themselves be modified). These pre-processing steps should be saved so that the complete path from raw data to results is transparent and reproducible, preferably in an R script in the "data-raw" directory. That script should end with creating an R data object in the "data" directory. Placing the R data object there ensures that it is included in the resulting R package, which makes using the data effortless for any R user, and eschews the need to download additional files. Loading the R package makes the included data immediately available in the R console.

Finally, the resulting data object in "data" should be documented in a standardized format by placing a `data.R` file in the "R" directory, a process discussed in more detail in our complete online supplementary tutorial (https://crumplab.github.io/vertical/), and in @WickhamPackagesOrganizeTest2015. That standard format of documentation is especially useful because it allows datasets' documentation to be viewed in R (e.g., type `?attitude` in the R console) and online (see below).

@@

The core of our proposal is that all project materials are stored in one directory on the user's local filesystem. When version control, such as Git, is enabled in this directory, the entire project can be seamlessly shared with others through online platforms such as GitHub, GitLab, and OSF. Version control systems provide filesystems with additional functioning for keeping track of files' versions, improved options for collaborating, and seamless integration with online collaboration platforms. For example, old versions of manuscripts are retained in the project's history. However, the details are outside our scope, and we refer readers to Vuorre & Curley [-@VuorreCuratingResearchAssets2018]. By default, *vertical* projects enable Git for new projects.

Because of the integration between Git(Hub) and one's local filesystem, users should primarily interact with Git(Hub). However, for long-term preservation, users should also link the project's GitHub repository to OSF. This can be done through the "Settings" page on an OSF project's page. Then, one can create an OSF registration of that project, which preserves the GitHub repository's current status on OSF permanently. This way, users gain both GitHub's convenience and OSF's long-term preservation.

@@

If the project contains R functions, data, and documentation following the R package standard, then others can install the package on their computers for fast access to the data and functions from R. Again, integrating the project with GitHub is beneficial, because then others can install the package from GitHub using the *devtools* package, as was shown above in how to install *vertical*. We recommend sharing most data packages through GitHub (as discussed here), rather than the official CRAN channel. Sharing as an R package is very convenient for other R users, as any packaged data and functions are immediately available for use, natively in R, after package installation. We expand on the extended benefits of using R packages for transparency and reproducibility in the general discussion.

## Reproducibility

An important component of computational reproducibility is specifying the (software) environment in which the computations were done, or should be done. R packages are a robust solution to this problem, because the package's metadata in the DESCRIPTION file can and should include the R packages and their versions that are required for the project. For example, if the project required using a specific version of R, and a specific version of an R package, this information could be written in the DESCRIPTION file.

@@

## Comparison to other approaches

We are not the first to recommend some form of standardization of workflows and procedures within psychology. First, Psych-DS is a collaborative project that aims to "promote the adoption of good practices in the management of scientific data" and "create a machine-readable format for these datasets that can support tools for analysis, discovery, and preparation of datasets in psychology" (https://github.com/psych-ds/psych-DS). Our suggestion is directly compatible with that effort, insofar as the R package standard doesn't require any special format from the raw data. In fact, future work might bring these two approaches together to inform both how the raw data should be formatted (Psych-DS), and how research products based on that raw data could best be curated and shared (present work).

There are also at least two other R packages designed to help researchers organize their workflows and facilitate reproducibility and data/code sharing. We highlight these packages here, while also discussing their similarities and differences to our approach. *workflowr* is an R packages that makes it easy to create version controlled and well-organized repositories of research materials, and share them as websites on GitHub [@blischakCreatingSharingReproducible2019; https://github.com/jdblischak/workflowr]. The *workflowr* R package itself is quite mature, and provides helpful tools for compiling many R Markdown documents sequentially, but has a slightly different approach to our proposal, in that workflowr does not itself provide much of a standardized organization for the materials (although it does provide a template). Further, the materials are not organized as an R package, as they would be in the (extended) *vertical* approach. Additionally, *workflowr* does not easily incorporate other types of materials (manuscripts, slides, posters, experiments) in the resulting website, unlike *vertical*. However, it may be slightly easier to use otherwise, for the many useful functions that it provides.

Another approach is contained in the *rrtools* R package [@marwickPackagingDataAnalytical2018; https://github.com/benmarwick/rrtools], which, unlike *workflowr* but like our approach, centers on the R package architecture. Additionally, *rrtools* helps set up a Docker container and continuous integration services for the project, both tools that we hope become more widely adopted in the future. However, it does not help users set up other important research materials as part of the project's materials, such as slides and experiments, nor does it help users share their projects as websites.

More generally, however, tools and descriptions, such as Psych-DS, *rrtools*, *vertical*, and *workflowr* are pointing to the same important and general idea: Psychological sciences would benefit from standardizing, as much as possible, the ways in which computational work is curated. We think that these tools and principles are in their infancy, and their importance is only now being recognized. Therefore, we expect these tools to evolve and mutate to best suit the practical researchers' needs, and hope that our suggestion provides some insight to how these tools might work in the future.

@@

Future work should also identify commonalities between our suggestion, and other related work such as rrtools [@marwickPackagingDataAnalytical2018] and workflowr [@blischakCreatingSharingReproducible2019], and potentially work to unify these approaches to a more comprehensible and user-friendly tool for research project organization.

@@

# Author Contributions

MV and MC jointly generated the idea for the manuscript; MV wrote the first draft, which both authors then critically edited. *vertical* was programmed by MC and MV. Both authors approved the final submitted version of the manuscript.
